Causal Inference in Population Trends: Searching for Demographic Anomalies in Big Data

### 2020-02-24 MANUSCRIPT/manuscript.Rmd
Updated. V1.0 of the manuscript.

abstract: "The proliferation of big data, wider access to advanced computing platforms, and the development of powerful statistical algorithms can uncover hidden anomalies in demographic data that have previously been dismissed as noise. Here, we demonstrate how social scientists can use causal inference techniques and abductive reasoning to identify fertility and mortality anomalies in state-level data trends by investigating demographic time series data since 1999. In demonstrating our technique, we uncover real spikes and dips in fertility and mortality over time, distinguished from regular trend variations. We show that 36 states exhibited at least one mortality anomaly over time, totaling nearly 300,000 additional deaths, while an additional seven states exhibited periods of decline resulting in 250,000 fewer deaths. Ten states exhibited real, marked periods of declines in fertility, totaling more than 214,000 fewer births than expected, and 8 states had periods of excess births, resulting in an addition 12,000 births. Our results show how researchers can detect demographic anomalies in trend data and use those anomalies as a beginning point for further investigation into the causes."

@@

The proliferation of data, advances in high performance (“super”) computing, and the development of powerful statistical algorithms mark the era of "Big Data" or data science [@van2016data;@zikopoulos2011], with the potential for researchers to find hidden or understudied social phenomena [@bohon2018demography]. While the availability of Big Data and high performance computing allows novel exploration of data through causal inference [@bohon2018demography;@rcausalimpact;@shiffrin2016drawing], relatively few studies utilize causal inference techniques in the study of demographic phenomena. However, understanding social phenomena using these advances reveals important insights into society [@angrist1989lifetime;@mas2009peers] and allows us to better monitor population trends [@nobles2019;@torche2015hidden].

Causal inference, in its most common usage, is an attempt to uncover the underlying mechanisms that result in changes in a phenomenon and has a long history as an approach to inquiry in the social sciences [@grimmer2015ppsp]. Often, the identification of a casual mechanism requires either a randomized control trial (RCT) or expert knowledge applied to a natural experiment. For population-level analysis, expensive RCTs usually preclude their widespread adoption [@west2008ajph]. Using natural experiments provide less certainty about causation and fewer opportunities to replicate work but they provide important and widespread insights.

@@

Both RCTs and natural experiments follow traditional hypothesis testing--inductive scientific paradigms where a question is first posed and scientists generate or find data to answer the question. Such approaches, while extensively utilized and time-tested, are likely to miss important phenomena that might go unnoticed. Data scientists are rethinking these approaches and have re-envisioned causal inference as a machine learning technique that allows big data researchers to move from correlations that have been established with good accuracy to making strong conclusions about the underlying mechanisms that produce these correlations
[@pearl2018book].

Abductive modeling is the movement from the inductive to the deductive, and sometimes back and forth, to reach conclusions [@bryant2014realm] or “inferring cause from effect” [@Crowder2017]. In situations with large data sets, an abductive approach is far superior to deductive hypothesis testing, as p-values with an extremely large number of cases are far from revealing [@head2015extent;@nuzzo2014scientific] and we do not want to make purely inductive inferences from data of questionable generalizability [@ruggles2014big]. Owing to the long history of big data sets in demographic research (one of the original "big data" resources [@ruggles2014big]), the rich demographic data available in the United States make the potential revelation of interesting and important demographic phenomena not only possible, but extremely plausible – even if identification of the phenomena occurs without identifying the underlying cause. In other sciences, the identification of effects without causes has led to new theories of galactic migration of planets and other breakthroughs [@gomes2005n].

In this paper, we use modern statistical outlier detection algorithms [@chen1993joint] as a means to illustrate one way that social scientists can engage in discovery using big data techniques on nearly twenty years of mortality and fertility data at the US state-level to identify anomalous demographic behavior. In essence, we identify *effects without knowledge of the cause*. We ask two questions regarding demographic anomalies: What are the hidden baby booms/busts and mortality spikes/dips in the United States over the last twenty years? We do not necessarily know the causes of these anomalies but identifying them allows scientists with more detailed knowledge of local population dynamics, state-level policy making, or macro-economics to explain these phenomena post-hoc. From explanations of phenomena that may have previously gone unnoticed, social scientists may be able to better forecast populations and provide policy solutions for impending problems such as climate change.

@@

# Results
We detect numerous anomalous mortality and fertility events at the US state-level since 1999. A full listing of these anomalies can be found in the **Supplementary Materials**. We begin with a summary of the anomalies we detect and then we highlight highly significant mortality and fertility anomalies across all three types of outliers (Additive Outliers, Level Shift Outliers, and Temporary Change Outliers) with plausible explanations. Finally, we highlight two strong anomalies that bely explanation.

@@

**\autoref{sumtable}** reports the overall number of anomalies we detect of each type for births and deaths and some summary statistics across all anomaly types and **\autoref{fig:summap}** maps these results.
We find considerably more mortality anomalies (n = `r paste(sums[2,2])`) than fertility anomalies (n = `r paste(sums[1,2])`). Given that anomalies are always the product of events [@song2018anomaly], finding more mortality than fertility anomalies is not surprising.  Mortality is likely to spike in response to a catastrophic event (like an earthquake or terrorist attack) or due to a disease outbreak while the effects of a catastrophic event on fertility is less predictable.  One reason for this is that fertility is linked to human decision-making more directly than mortality [@stein2014couples] which results in more varied outcomes, so, for example, researchers find that catastrophic weather events can increase childbearing among those who already have a child, but not among the childless [@Evans2008Hurricanebirth]. Some events, like the 1995 Oklahoma City bombings, resulted in both fertility and mortality changes.  However, changes in fertility manifest over a longer time horizon after an event, and the relationship between a fertility-inducing event and behavior change is less strong than the relationship between a mortality-inducing event (such as a terrorist incident) and death [@Rodgers2005OKBombing]. Of course, not all events that impact fertility and mortality are catastrophic. Researchers have documented that more commonplace events such as massive layoffs [@Venkataramani2019] or policy changes [@Livingston2017Cannabis] can also create anomalies in mortality patterns, but we have not found such evidence for fertility.

@@

## Mortality 
Since the purpose of our paper is to demonstrate how social scientists can make use of causal inference technique, we will not discuss all of our findings in this paper.  Instead, we will use a few cases for illustration.  For mortality, we will use New York State as an example (**\autoref{fig:mortnewhamp}a**). We identify seven anomalies in the mortality time series for New York, all with *t*-statistics in excess of 3.91, making these anomalies unlikely to be due to chance. The algorithm correctly identifies September 2001 as an additive outlier (2001:09 *t*= `r paste(prettyNum(outlier.mort.ny$outliers[2,5], digits = 4))`) where there were `r paste(prettyNum(outlier.mort.ny$outliers$coefhat[2], digits = 4, big.mark = ","))` more deaths in that month than anticipated. This mortality event is likely caused by the September 11 terrorist attack on the World Trade Center that immediately killed 2,606 people and the detection of this mortality event provides confidence in our detection of other anomalies.

```{r MortalityAnomalies, echo= FALSE, message = FALSE, warning = FALSE,dev=c('pdf','png'), fig.height = 5, fig.cap= paste("\\textbf{Anomaly Detection for New York (a) and New Hampshire (b) state mortality, 1999-2016.} The top part of each panel contains the original time series (light gray), the corrected, counter-factual time series in the absence of anomalies (blue), and the red dots correspond to the onset of detected anomalies. The bottom part of each panel contains the magnitude and type of the outlier in red. In New York (a), we detect additive outliers (AO) in January 2000, September 2001, January 2005, and January 2013; temporary change (TC) outliers in February 2007 and January 2015; and a level shift (LS) starting in February 2004. In New Hampshire (b), we detect two outliers, both level shift outliers (LS) in April 2010 and again in November 2014. These anomalies suggest New York experienced a significant mortality event in September 2001 and New Hampshire experienced approximately 9,700 more deaths than expected since 2010 or 14% more deaths in the state over just seven years. \\label{fig:mortnewhamp}")}

To illustrate how researchers can examine fertility anomalies, we begin with a clear temporary change (TC) outlier for fertility in Louisiana (**\autoref{fig:fertla}a**). Here we detect two TC outliers, nearly back-to-back in 2005 (2005:08 *t* = `r la_tstat1`; and 2005:10 *t* = `r la_tstat2`), coinciding with the destruction of Hurricane Katrina from the same year. The migration associated with Hurricane Katrina has garnered most of the attention of social scientists [@fussellRecoveryMigrationCity2014;@horiDisplacementDynamicsSouthern2009], but the hurricane had a very clear impact on fertility behaviors too, creating a mini "baby bust" in Louisiana with the departure of many people in their childbearing years. We estimate `r la_averbirths` fewer births in Louisiana compared to the counterfactual, likely attributable to the Hurricane.

```{r FertilityAnomalies, echo= FALSE, message = FALSE, warning = FALSE, fig.cap= paste("\\textbf{Anomaly detection in Louisiana (a) and Connecticut (b) state fertility, 2003-2018.} In (a), we detect two outliers, back-to-back, likely resulting from Hurricane Katrina in August and October 2005, representing a decrease of more than 4,400 births due to the hurricane. In (b), we detect one outlier, a level shift outlier (LS) in August 2009. We believe this reduction is attributable to the stock market crash 11 months earlier. \\label{fig:fertla}")}

# Conclusion
Data scientists frequently claim that the big data revolution is a turning point in scientific discovery that will allow us to solve some of the world’s most pressing problems [@grimmer2015ppsp].  Social scientists are skeptical of such claims, because they better understand the complexities of the social world and know from experience that data, alone, is not enough [@bohon2018demography;@grimmer2015ppsp]. Nonetheless, the increased availability of data and (more importantly, we argue) the development of advanced techniques for analyzing these data will enable important discovery [@monroe2015no].

>>>>

# abstract: "The proliferation of big data, wider access to advanced computing platforms, and the development of powerful statistical algorithms can uncover hidden anomalies in demographic data that have previously been dismissed as noise. Here, we demonstrate how social scientists can use causal inference techniques and abductive reasoning to identify fertility and mortality anomalies in state-level data trends by investigating demographic time series data since 1999. In demonstrating our technique, we uncover real spikes and dips in fertility and mortality over time, distinguished from regular trend variations. We show that 36 states exhibited at least one mortality anomaly over time, totaling nearly 300,000 additional deaths, while an additional seven states exhibited periods of decline resulting in 250,000 fewer deaths. Ten states exhibited real, marked periods of declines in fertility, totaling more than 214,000 fewer births than expected, and 8 states had periods of excess births, resulting in an addition 12,000 births. Our results show how researchers can detect demographic anomalies in trend data and use those anomalies as a beginning point for further investigation into the causes."

abstract: "The proliferation of big data, wider access to advanced computing platforms, and the development of powerful statistical algorithms can uncover hidden anomalies in social data, previously dismissed as noise. Here, we combine causal inference techniques and abductive reasoning to identify fertility and mortality anomalies on twenty years of demographic data in the United States. We uncover real, “hidden” baby booms/busts and mortality spikes/dips, distinguishable from regular trend variations.  We identify more than 22 and 156 fertility and mortality anomalies, totaling more than 200k and 600k anomalous births and deaths, respectively. Notable detectable mortality anomalies include the September 11 2001 terrorist attack in New York, the emergence and acceleration of the opioid epidemic in New Hampshire, and the excess mortality due to Hurricane Maria in Puerto Rico. Notable fertility anomalies include the “missing births” in Louisiana after Hurricane Katrina and the reduction in fertility behavior after the September 2008 stock market crash in Connecticut, amongst others. Our combined causal inference and abductive reasoning approach can be readily adapted to find other, undiscovered social phenomena or to evaluate the efficacy of important public policies."

@@

The search for causality is one of the primary motiviations behind social science [@smith2011commentary; @pearl2009causal]. Causal Inference, the identification of causal mechanisms, attempts to uncover the underlying mechanisms that result in changes in a phenomenon and has a long history as an approach to inquiry in the social sciences [@grimmer2015ppsp]. Often, the identification of a casual mechanism is accomplished via a randomized control trial (RCT) or expert knowledge applied to a natural experiment [@salganik2019bit].

Unfortunately, both RCTs and natural experiments have drawbacks. RCTs are often prehibitively expensive for many scientists, precluding their widespread adoption in the social sciences [@west2008ajph]. RCTs are also sometimes impossible to deploy in the "real world". For instance, an RCT to study country-level austeriy measures would be difficult, if not impossible, to implement. Natural experiments are a cheaper alternative, but provide less certainty about causation, fewer opportunities to replicate work, and require expert knowledge of the natural experiment to be properly employed [@pearl2018book].

<!-- In its most basic application, causal inference approaches accomplish this with two populations, a control group ($Y(0)$) and a treatment group ($Y(1)$) and simply subtracting the results of the control from the treatment ($effect = Y(1) - Y(0))$).  -->

@@

Additionally, both RCTs and natural experiments follow traditional hypothesis testing, the deductive scientific paradigm where a question is first posed and scientists generate or find data to answer the question. Such approaches, while extensively utilized and time-tested, are likely to miss important phenomena that might go unnoticed or suffer from extensive p-hacking [@head2015extent;@nuzzo2014scientific;@ruggles2014big]. Abductive reasoning, sometimes described as “inferring cause from effect” [@Crowder2017], moves the scientific process from the inductive to the deductive, and sometimes back and forth, to reach conclusions [@bryant2014realm] The physical sciences use abductive reasoning more fully compared to the social sciences. The discovery of Neptune in 1846 based on unexplained perturbations of the orbit of Uranus [@popper2005logic], the discovery of galactic migration of planets [@gomes2005n], and the current generation of gravitational wave detectors such as the Laser Interferometer Gravitational-Wave Observatory (LIGO) [@harry2010advanced] demonstrate the power of the abductive approach in the physical sciences.

In a data-rich environment, abductive reasoning can be more easily applied to first find effects and then infer their cause ex-post. The modern proliferation of data, advances in high performance (“super”) computing, and the development of powerful statistical algorithms provide an environment ripe for wedding causal inference with abductive reasoning [@van2016data;@zikopoulos2011], with the potential for researchers to find hidden or understudied social phenomena [@bohon2018demography]. While the availability of Big Data and high performance computing allows novel exploration of data through causal inference [@bohon2018demography;@rcausalimpact;@shiffrin2016drawing], relatively few studies utilize causal inference techniques in the study of demographic phenomena. However, understanding social phenomena using these advances reveals important insights into society [@angrist1989lifetime;@mas2009peers] and allows us to better monitor population trends [@nobles2019;@torche2015hidden]. The rich demographic data available in the United States make the potential revelation of interesting and important demographic phenomena not only possible, but extremely plausible – even if identification of the phenomena occurs without identifying the underlying cause.

<!-- Data scientists are rethinking these approaches and have re-envisioned causal inference as a machine learning technique that allows big data researchers to move from correlations that have been established with good accuracy to making strong conclusions about the underlying mechanisms that produce these correlations [@pearl2018book]. The discovery of Neptune in 1846 In other sciences, the identification of effects without causes has led to new theories of galactic migration of planets and other breakthroughs [@gomes2005n]. -->

<!-- Abductive modeling is the movement from the inductive to the deductive, and sometimes back and forth, to reach conclusions [@bryant2014realm] or “inferring cause from effect” [@Crowder2017]. In situations with large data sets, an abductive approach is far superior to deductive hypothesis testing, as p-values with an extremely large number of cases are far from revealing [@head2015extent;@nuzzo2014scientific] and we do not want to make purely inductive inferences from data of questionable generalizability [@ruggles2014big]. Owing to the long history of big data sets in demographic research (one of the original "big data" resources [@ruggles2014big]),  -->

In this paper, we combine a modern statistical outlier detection algorithm [@chen1993joint], abductive reasoning, and large demographic data sets on more than 20 years of mortality/fertility to identify anomalous demographic behavior. In essence, we identify *effects without knowledge of the cause*. We ask one fundamanetal question regarding demographic anomalies: What are the hidden baby booms/busts and mortality spikes/dips in the United States over the last twenty years? We do not necessarily know the causes of these anomalies but identifying them allows scientists with more detailed knowledge of local population dynamics, state-level policy making, or macro-economics to explain these phenomena post-hoc. From explanations of phenomena that may have previously gone unnoticed, social scientists may be able to better forecast populations and provide policy solutions for impending problems such as climate change.

@@

# Results
We detect numerous anomalous mortality and fertility events at the US state-level since 1999. A full listing of these anomalies can be found in the **Supplementary Materials**. We begin by highlighting significant fertility and mortality anomaloies across all three types of outliers (Additive Outliers, Level Shift Outliers, and Temporary Change outliers) with plausible explanations. We then highlight two strong anomalies that bely explanation. Finally, we conclude with a summary of the anomalies we detect.

We begin with a clear temporary change (TC) outlier for fertility in Louisiana (**\autoref{fig:fertla}a**). Here we detect two TC outliers, nearly back-to-back in 2005 (2005:08 *t* = `r la_tstat1`; and 2005:10 *t* = `r la_tstat2`), coinciding with the destruction of Hurricane Katrina from the same year. The migration associated with Hurricane Katrina has garnered most of the attention of social scientists [@fussellRecoveryMigrationCity2014;@horiDisplacementDynamicsSouthern2009], but the hurricane had a very clear impact on fertility behaviors too, creating a mini "baby bust" in Louisiana with the departure of many people in their childbearing years. We estimate `r la_averbirths` fewer births in Louisiana compared to the counterfactual, likely attributable to the Hurricane.

```{r FertilityAnomalies, echo= FALSE, message = FALSE, warning = FALSE, fig.cap= paste("\\textbf{Anomaly detection in Louisiana (a) and Connecticut (b) state fertility, 2003-2018.} The top part of each panel contains the original time series (light gray), the corrected, counter-factual time series in the absence of anomalies (blue), and the red dots correspond to the onset of detected anomalies. The bottom part of each panel contains the magnitude and type of the outlier in red. In (a), we detect two outliers, back-to-back, likely resulting from Hurricane Katrina in August and October 2005, representing a decrease of more than 4,400 births due to the hurricane (t-statistics= -6.002 and -4.997). In (b), we detect one outlier, a level shift outlier (LS) in August 2009. We believe this reduction is attributable to the stock market crash 11 months earlier (t-statistic = -3.71). \\label{fig:fertla}")}

## Mortality 
For mortality, we will use New York State as an example (**\autoref{fig:mortnewhamp}a**). We identify seven anomalies in the mortality time series for New York, all with *t*-statistics in excess of 3.91, making these anomalies unlikely to be due to chance. The algorithm correctly identifies September 2001 as an additive outlier (2001:09 *t*= `r paste(prettyNum(outlier.mort.ny$outliers[2,5], digits = 4))`) where there were `r paste(prettyNum(outlier.mort.ny$outliers$coefhat[2], digits = 4, big.mark = ","))` more deaths in that month than anticipated. This mortality event is likely caused by the September 11 terrorist attack on the World Trade Center that immediately killed 2,606 people and the detection of this mortality event provides confidence in our detection of other anomalies.

```{r MortalityAnomalies, echo= FALSE, message = FALSE, warning = FALSE,dev=c('pdf','png'), fig.height = 5, fig.cap= paste("\\textbf{Anomaly Detection for New York (a) and New Hampshire (b) state mortality, 1999-2016.} In New York (a), we detect additive outliers (AO) in January 2000, September 2001, January 2005, and January 2013; temporary change (TC) outliers in February 2007 and January 2015; and a level shift (LS) starting in February 2004. In New Hampshire (b), we detect two outliers, both level shift outliers (LS) in April 2010 and again in November 2014. These anomalies suggest New York experienced a significant mortality event in September 2001 and New Hampshire experienced approximately 9,700 more deaths than expected since 2010 or 14% more deaths in the state over just seven years. \\label{fig:mortnewhamp}")}

@@

**\autoref{sumtable}** reports the overall number of anomalies we detect of each type for births and deaths and some summary statistics across all anomaly types and **\autoref{fig:summap}** maps these results.

We find considerably more mortality anomalies (n = `r paste(sums[2,2])`) than fertility anomalies (n = `r paste(sums[1,2])`). Given that anomalies are always the product of events [@song2018anomaly], finding more mortality than fertility anomalies is not surprising.  Mortality is likely to spike in response to a catastrophic event (like an earthquake or terrorist attack) or due to a disease outbreak while the effects of a catastrophic event on fertility is less predictable.  One reason for this is that fertility is linked to human decision-making more directly than mortality [@stein2014couples] which results in more varied outcomes, so, for example, researchers find that catastrophic weather events can increase childbearing among those who already have a child, but not among the childless [@Evans2008Hurricanebirth]. Some events, like the 1995 Oklahoma City bombings, resulted in both fertility and mortality changes.  However, changes in fertility manifest over a longer time horizon after an event, and the relationship between a fertility-inducing event and behavior change is less strong than the relationship between a mortality-inducing event (such as a terrorist incident) and death [@Rodgers2005OKBombing]. Of course, not all events that impact fertility and mortality are catastrophic. Researchers have documented that more commonplace events such as massive layoffs [@Venkataramani2019] or policy changes [@Livingston2017Cannabis] can also create anomalies in mortality patterns, but we have not found such evidence for fertility.

@@

# Conclusion
Data scientists frequently claim that the big data revolution is a turning point in scientific discovery that will allow us to solve some of the world’s most pressing problems [@grimmer2015ppsp].  Social scientists are skeptical of such claims, because they better understand the complexities of the social world and know from experience that data alone is not enough to solve social problems [@bohon2018demography;@grimmer2015ppsp]. Nonetheless, the increased availability of data and (more importantly, we argue) the development of advanced techniques for analyzing these data will enable important discovery [@monroe2015no].
